---
title: "MY472 - Week 3: ggplot2 basics - axes, scales, legends, polishing plots"
author: "Pablo Barberá, Friedrich Geiecke"
date: "AT 2024"
output: html_document
---

## Introduction

We will continue working with a random sample of 10,000 public Facebook posts by members of the U.S. Congress. This file focuses on axes, legends, scales, facets, and polishing plots more generally.

For a review of the log scale, see for example this [video](https://www.youtube.com/watch?v=sBhEi4L91Sg).

```{r}
# install.packages("gridExtra") ## IF NOT INSTALLED YET
library("tidyverse")
library("scales")
library("gridExtra")
```

```{r}
df <- read.csv("data/fb-congress-data.csv", stringsAsFactors = TRUE)

# Transform date column to datetime
df$date <- as_date(df$date)

# Dropping some very uncommon post types
nrow(df)
df <- df %>%
  filter(!post_type %in% c("music", "note"))
nrow(df)
```

## Customizing axes

```{r}

# Changing axis titles
p <- ggplot(data = df, mapping = aes(x = likes_count))
p + geom_histogram() + scale_x_log10("Number of likes", labels = comma) +
    scale_y_continuous("Post count", labels = comma)

# Changing axis limits (continuous variables)
p + geom_histogram() + scale_y_continuous("Post count") +
  scale_x_log10("Number of likes", limits = c(1, 100000), labels = comma)
    
# Changing axis breaks and defining label for each break
p <- ggplot(df, aes(x = likes_count)) + xlab("likes count")
p + geom_histogram() + scale_x_log10(breaks = c(1, 10, 100, 1000))
p + geom_histogram() + scale_x_log10(breaks = c(1, 10, 100, 1000, 10000),
    labels=c(1, 10, 100, "1K", "10K"))

# Changing axis breaks and labels (categorical variables)
p <- ggplot(df, aes(x=party))
p + geom_bar() + scale_x_discrete(labels=c("D", "I", "R"))

# Creating data for time series plot of posts per day
counts <- df %>%
  group_by(date) %>% 
  summarise(posts = n())

# Basic line plot
p <- ggplot(counts, aes(x = date, y = posts))
p + geom_line()

# Intervals of two months
p + geom_line() + scale_x_date(date_breaks = "2 months")

# Using the scales package to only depict year and month
p + geom_line() + scale_x_date(date_breaks = "2 months", 
    labels = date_format("%Y-%m"))
```

## Customizing scales and legends

```{r}
p <- ggplot(df, aes(x = likes_count)) + xlab("likes count")
p + scale_x_log10(labels = comma) + geom_histogram(color = "red") # border of geoms
p + scale_x_log10(labels = comma) + geom_histogram(fill = "red") # area of geoms

p <- ggplot(df, aes(x = likes_count, y = comments_count)) +
  scale_x_log10("likes count", labels = comma) +
  scale_y_log10("comments count", labels = comma)

p + geom_point(color = "red")
p + geom_point(shape = 15)
p + geom_point(size = 1)

# for a complete list for each geom, type ?geom_point, ?geom_line, ...
```

Next, let us use color within aesthetics and in combination with the post type. Rather than as a property of the geom, it now applies to specific groups (here post types) via scales  instead of to all data points (when specified in geom before).

```{r}
p <- ggplot(df, aes(x = likes_count, y = comments_count)) +
  scale_x_log10("Likes count", labels = comma) +
  scale_y_log10("Comments count", labels = comma)
p + geom_point(aes(color = post_type)) +
    scale_color_manual("Post type",
        limits = c("link", "status", "video", "photo"),
        values = c("blue", "grey", "red", "yellow"))

p + geom_point(aes(shape = post_type)) +
    scale_shape_manual("Post type",
        limits = c("link", "status", "video", "photo"),
        values = c(1, 3, 4, 5)) # integers specify shapes
```

Finally, we can also modify the labels in the legend:

```{r}
p + geom_point(aes(shape = post_type)) +
    scale_shape_manual("Post type",
        limits = c("link", "status", "video", "photo"),
        labels = c("Link", "Status", "Video", "Photo"),
        values=c(1, 3, 4, 5))
```

## Facets

Facets can e.g. be very helpful when we would like to repeat plots for different values of a categorical variable. Here we benefit from data in a tidy format where classes are stored in a distinct variable.

```{r}
p + geom_point() + facet_wrap(~post_type)

p + geom_point() + facet_wrap(~post_type, nrow=4)

p + geom_point() + facet_wrap(~post_type, ncol=4)
```

With two factor variables:

```{r}
p + geom_point() + facet_grid(gender~post_type)
```

Combining facets and scales:

```{r}
p <- ggplot(df[df$likes_count>5000,], 
            aes(x = likes_count, y = comments_count)) +
  scale_x_log10("Likes count", labels = comma) +
  scale_y_log10("Comments count", labels = comma)
p + geom_point(aes(shape = gender, color = post_type, size = shares_count)) +
  scale_size(trans = "log10") + facet_wrap(~post_type, nrow = 2)  
```


## Changing themes

```{r}
p + geom_point() + theme_bw()

p + geom_point() + theme_grey()

p + geom_point() + theme_minimal()

p + geom_point() + theme_classic()
```

All theme options can be edited manually using 'theme', e.g. removing axis ticks:

```{r}
p + geom_point() + theme( axis.ticks = element_blank() )
```

See `?theme` for all possible options. We will see more examples later.

## Saving plots

Different ways of doing this, easiest is with `ggsave`:

```{r}
pq <- p + geom_point() + facet_grid(party~gender)

ggsave(pq, file="grid_plot.pdf", height=6, width=6)
```

## Multiple types of plots on a grid

Now we will combine multiple plots on a grid. For this we use the `gridExtra` package which adds more functions.

```{r}
# Posts by type of post
p1 <- ggplot(df, aes(x = post_type)) + geom_bar() + theme_minimal() +
    scale_y_continuous("Number of posts") + 
    scale_x_discrete("Type of post") +
    theme (axis.ticks.x = element_blank())

p2 <- ggplot(df, aes(x = gender)) + geom_bar() + theme_minimal() +
     scale_x_discrete("Gender of legislator", 
        labels=c("Female", "Male")) +   
    theme (axis.title.y = element_blank(), axis.text.y = element_blank(),
        axis.ticks = element_blank())

# average number of likes by type of post
p3 <- ggplot(df, aes(x = post_type, y = likes_count)) + 
    stat_summary(fun = "mean", geom = "point", size = 5, shape = 15) +
    theme_minimal() + scale_y_continuous("Average likes count") +
    scale_x_discrete("Type of post") +
    theme (axis.ticks.x = element_blank()) 

# average number of likes by gender
p4 <- ggplot(df, aes(x = gender, y = likes_count)) + 
    stat_summary(fun = "mean", geom = "point", size = 5, shape = 15) +
    theme_minimal() + scale_x_discrete("Gender of legislator", 
        labels = c("Female", "Male")) +   
    theme (axis.title.y = element_blank(), axis.text.y = element_blank(),
        axis.ticks = element_blank())


# dev.off() closes the file
pdf("multiple_plots_grid.pdf", height = 6, width = 6)
grid.arrange(arrangeGrob(p1, p2, p3, p4, ncol = 2, heights = c(0.6, 0.4)))
dev.off()
```
