---
title: "Reading JSON into R"
author: "Friedrich Geiecke"
date: "26/10/2020"
output: html_document
---

Loading packages:

```{r}
library("jsonlite")
library("tidyverse")
```

Specify the directory where the JSON files are stored:

```{r}
current_dir <- "yourdirectoryhere"
```

Reading/parsing the first file into R:

```{r}
example_1_parsed <- fromJSON(paste(current_dir, "example_1.json", sep = ""))
class(example_1_parsed)
names(example_1_parsed)
```

The key-value structure is of JSON objects is transformed into (nested) lists in R:

```{r}
example_1_parsed
```

We can e.g. access the grades array with:

```{r}
example_1_parsed$grades
```

The JSON array became a vector in R. Also note e.g. that the JSON notation of the Boolean as `false` was transformed into the R notation of logical `FALSE`.

The supervisor element is again a list itself. This emphasises the nested structure of such files:

```{r}
class(example_1_parsed$supervisor)
```

We can access the supervisor's department with:

```{r}
example_1_parsed$supervisor$department
```

Let us proceed with the second example. Here we had built the JSON file such that it contained a tabular data structure. We can parse the file in the usual list format, however, also transform it into a dataframe.

```{r}
example_2_parsed <- fromJSON(paste(current_dir, "example_2.json", sep = ""))
class(example_2_parsed)
names(example_2_parsed)
```

List:

```{r}
example_2_parsed
```

Dataframe:

```{r}
example_2_parsed %>% as_tibble()
```

Note that the JSON `null` was transformed into an NA.

Lastly, let us look at a slightly more realistic example. Again we obtain a list for the JSON file, however, `jsonlite` actually understands here that each of the keys in the JSON (MT and LT) stores tabular data which can be represented as a dataframe in R:

```{r}
example_3_parsed <- fromJSON(paste(current_dir, "example_3.json", sep = ""))
class(example_3_parsed)
names(example_3_parsed)
```

The dataframe contained in the MT key:

```{r}
example_3_parsed$MT
```

The dataframe contained in the LT key:

```{r}
example_3_parsed$LT
```

