---
title: "MY472 - Week 5: Optional advanced exercise for scraping unstructured data - solution"
author: "Friedrich Geiecke"
date: "27 October 2022"
output: html_document
---

This notebook studies an advanced example of scraping unstructured data. The task is to collect information on Nobel laureates in physics via Wikipedia. This is done by navigating to the page of each laureate and then collecting information from the content of that page. This logic, i.e. navigating from a base page that contains links to a range of other pages, is widely applicable. Think e.g. of a website with names of firms, countries, politicians, etc. each linking to an own page that contains information about the respective entry.

Note: While Wikimedia also has the Wikidata API which would allow to obtain data in an already structured format here, the Wikipedia website itself is a great training ground to study web scraping.

```{r}
library("tidyverse")
library("rvest")
```

## 1. Extracting information from a Wikipedia page

Let us look at Richard Feynman's Wikipedia page as a single example first. As a side note, for an interesting discussion of the scientific method by Feynman, see this [link](https://youtu.be/EYPapE-3FRw); in particular the first 2.5 minutes are also relevant for social scientists that work with mathematical models.

```{r}
url_feynman <- "https://en.wikipedia.org/wiki/Richard_Feynman"
html_feynman <- read_html(url_feynman)
```

Say we would like to collect some baseline information about the person. The box on the right hand side seems the natural starting point. Each item in this box has content in a cell on the left (e.g. "Alma mater") and on the right (e.g. "Massachusetts Institute of Technology (S.B.) Princeton University (PhD)"). Inspecting with the browser reveals that the left cell has a class name "infobox-label" and the right "infobox-data". Try to complete the following by adding the CSS selectors for these two class names. This allows to obtain all relevant label and data elements from the box:

```{r}
label_elements_feynman <- html_feynman %>% html_elements(css = ".infobox-label")
data_elements_feynman <- html_feynman %>% html_elements(css = ".infobox-data")
```

Next, extract the text from these elements:

```{r}
labels_feynman <- label_elements_feynman %>% html_text()
data_feynman <- data_elements_feynman %>% html_text()
```

Printing the first entries:

```{r}
labels_feynman[1:5]
```

```{r}
data_feynman[1:5]
```
From these vectors, how would you access just the birth date information?

```{r}
data_feynman[labels_feynman=="Born"]
```

The title of his PhD thesis?

```{r}
data_feynman[labels_feynman=="Thesis"]
```

The information could now be stored in a data frame. In the next section, we will look at how we can automate this approach for all physics laureates rather than just Richard Feynman.

## 2. Combining information from multiple pages

The following Wikipedia page contains all Nobel laureates' names and links to individual pages:

```{r}
url <- "https://en.wikipedia.org/wiki/List_of_Nobel_laureates"
nobels_html <- read_html(url) # reading the HTML code
```

You can scrape the full table by adding its CSS selector here:

```{r}
nobel_table_wikipedia <- html_elements(nobels_html, css = ".wikitable")[[1]] %>% html_table(fill=TRUE)
nobel_table_wikipedia
```

This does return the names of all laureates, however, no further information about them such as their institutions, dissertation supervisors, etc. For the example of only the physics laureates, let us therefore obtain such data.

First, create vector containing only the physics laureates with one element being one scientist:

```{r}
# Vector with all names of physics laureates (requires to separate multiple laureates in some years)
physicists_names <- nobel_table_wikipedia$Physics %>% str_split(";") %>% unlist()

# Drop entries of the physics column which don't contain names (years without laureates)
physicists_names <- physicists_names[physicists_names != "None"]
physicists_names <- physicists_names[physicists_names != "Cancelled due to World War II"]

# Drop the last row of the column which just says "Physics"
physicists_names <- physicists_names[-length(physicists_names)]

# Exemplary names
length(physicists_names)
physicists_names[1:10]
```

The information later collected for each laureate will be stored in a table with columns for fields, institutions, doctoral advisors, and doctoral students.

```{r}
physicists <- tibble(name = physicists_names, fields = NA, institutions = NA, doctoral_advisors = NA, doctoral_students = NA)
physicists
```

Next, store all hyperlinks contained in the website's HTML code in a vector (what is the CSS selector for a hyperlink?). This is one possible approach which allows to navigate to each individual laureate's page later:

```{r}
all_link_elements <- html_elements(nobels_html, css = "a")
all_link_texts <- all_link_elements %>% html_text()
```

In the following, some code to extract and clean texts from a person's Wikipedia page is wrapped into a helper function. It uses the same idea as the Feynman example considered initially and furthermore cleans the texts. It works relatively well to clean texts for fields, institutions, doctoral advisors, and doctoral students as these are needed to fill out the table above. Other entries, such as the birth year, would require cleaning with a different set of regular expressions.

```{r}
# Helper function to clean texts
helper_function <- function(x, labels, data) {
  
  #
  # Only works well with fields, institutions, doctoral advisors, and doctoral students
  #
  
  # First, also create a plural version of the label string, e.g. if the label
  # "x" is "advisor", the function will also check whether the Wikipedia page
  # mentions "advisors" instead
  x_plural <- paste0(x,"s")
  
  # Check singular
  if (x %in% labels) {
    text <- data[labels==x]
  }
  # Check plural
  else if (x_plural %in% labels) {
    text <- data[labels==x_plural]
    }
  else {return(NA)} # return statement stops the rest of the function from running
  
  ## Clean text
  # Remove new line signs and replace them with ;
  text <- text %>% str_replace_all("\n", "; ")
  # Remove any resulting ; at the beginning of the string
  text <- text %>% str_replace_all("^; ", "")
  # Remove superscripts and content added in parentheses
  text <- text %>% str_replace_all("\\[.*\\]", "")
  text <- text %>% str_replace_all("\\(.*\\)", "")
  # Sometimes words in the Wikipedia texts are not correctly separated with a white space
  # The following e.g. transforms "University of LondonCambridge University" into
  # "University of London; Cambridge University"
  text <- text %>% str_replace_all("([[:lower:]])([[:upper:]])", "\\1; \\2") # note: [[:lower:]] e.g. contains é which is not contained in [a-z]
  # Sometimes long residual website code is contained at the beginning of text before
  # the first ; -> It will be deleted in the following
  text_split <- str_split(text, "; ")[[1]]
  if (nchar(text_split[1]) > 200) { # Before the first semicolon, are there more than 200 characters?
    text <- text_split[2:length(text_split)] %>% paste(collapse = "; ")}
  return(text)
  
}

## Illustration with the Feynman examples discussed earlier

# Obtain institution
helper_function("Institutions", labels_feynman, data_feynman)
# Obtain doctoral advisor(s)
helper_function("Doctoral advisor", labels_feynman, data_feynman)
```

Putting all these pieces together allows to collect the information for each physics laureate by navigating to their individual page:

```{r}
# Loop over all names of physics laureates
for (current_name in physicists_names) {
  
  # Obtain relevant link element for the current laureate's name
  current_laureate_element <- all_link_elements[all_link_texts == current_name][1]
  
  ## Navigate to URL of current laureate
  # Get target of hyperlink
  current_partial_url <- current_laureate_element %>% html_attr("href")
  current_url <- paste("https://en.wikipedia.org",  current_partial_url, sep = "")
  # Load page
  current_html <- read_html(current_url)
  
  # Get labels and data
  labels <- current_html %>% html_elements(css = ".infobox-label") %>% html_text()
  data <- current_html %>% html_elements(css = ".infobox-data") %>% html_text()
  
  # Clean text and store in data frame
  physicists[physicists$name == current_name, "fields"] <- helper_function("Field", labels, data)
  physicists[physicists$name == current_name, "institutions"] <- helper_function("Institution", labels, data)
  physicists[physicists$name == current_name, "doctoral_advisors"] <- helper_function("Doctoral advisor", labels, data)
  physicists[physicists$name == current_name, "doctoral_students"] <- helper_function("Doctoral students", labels, data)
  
  # Wait half a second before next request
  Sys.sleep(0.5)
  
}

physicists
```

The cleaned texts in the data frame are not 100% consistent and some further fine tuning of the regular expressions would be required for rare cases, however, the data quality is already quite good. The example shows that with some lines of code and navigating through Wikipedia it was already possible to assemble a notable database of scientists, their institutions, advisors and doctoral students all from different webpages.

Who was the doctoral advisor of most scientists that became physics prize winners?

```{r}
all_advisors <- physicists$doctoral_advisors %>% str_split("; ") %>% unlist()
all_advisors <- tibble(advisor = all_advisors) %>% drop_na()
all_advisors %>% count(advisor) %>% arrange(desc(n))
```
