---
title: "MY472 - Week 11: Seminar exercises on BigQuery - solution"
date: "Autumn Term 2024"
output: html_document
---

**Attribution statement:** _The following teaching materials have been iteratively developed by current and former instructors, including: Pablo Barberá, Friedrich Geiecke, Daniel de Kadt and Ryan Hübert._

## Data

We will use the NYC Bicycle Hire data we looked at in lecture. The database contains Citi Bike trips (NYC's bike-sharing service) since Citi Bike launched in September 2013.

```{r}
library("DBI")
library("bigrquery")
```

## Queries

1. First, connect to the database and see the first 5 entries.

```{r}
# Project name (now accessing the public datasets)
project_name <- "bigquery-public-data"

# Billing (main project ID)
billing_info <- "tba"

# Dataset name (part of the public datasets)
dataset_name <- "new_york"
```

```{r}
db <- dbConnect(
  bigrquery::bigquery(),
  project = project_name,
  dataset = dataset_name,
  billing = billing_info
)
db
```

All tables in the database:

```{r}
dbListTables(db)
```

```{r}
# First 5 entries
dbGetQuery(db,
           "SELECT *
           FROM citibike_trips
           LIMIT 5")
```

2. What is the average trip duration based on the age of the riders?

```{r}
res2 <- dbGetQuery(db,
                   "SELECT (2020-birth_year) AS age,
                      AVG( TIMESTAMP_DIFF(stoptime, starttime, MINUTE) ) AS duration_minutes,
                      COUNT(*) AS trip_count
                   FROM citibike_trips
                   GROUP BY age
                   ORDER BY age")
```

```{r}
# Age against duration
ggplot(data = subset(res2, age < 100), aes(x = age, y = duration_minutes)) +
  geom_line() +
  labs(x = "Age", y = "Duration (minutes)",
       title = "Age vs. Duration Minutes") +
  theme_minimal()
# Age distribution
ggplot(data = subset(res2, !is.na(age) & age < 100), aes(x = age, y = trip_count)) +
  geom_bar(stat = "identity") +
  labs(x = "Age", y = "Trip Count",
       title = "Age Distribution") +
  theme_minimal()
```

3. What is the average distance of a CitiBike trip based on the age of the riders?

```{r}
res3 <- dbGetQuery(db,
                   "SELECT (2017-birth_year) AS age,
                      AVG( (ABS(start_station_latitude-end_station_latitude) +
                            ABS(start_station_longitude-end_station_longitude)) * 111) AS avg_distance_km 
                   FROM citibike_trips
                   GROUP BY age
                   ORDER BY age")
```

```{r}
ggplot(data = subset(res3, age < 100), aes(x = age, y = avg_distance_km)) +
  geom_line() +
  labs(x = "Age", y = "Average Distance (km)",
       title = "Age vs. Average Distance (km)") +
  theme_minimal()
```

4. What is the average speed of a CitiBike trip?

```{r}
dbGetQuery(db,
          "SELECT AVG( ((ABS(start_station_latitude-end_station_latitude) + 
                        ABS(start_station_longitude-end_station_longitude)) * 111) /
                    (TIMESTAMP_DIFF(stoptime, starttime, MINUTE)/60) ) AS avg_speed_km_h
          FROM citibike_trips")
```

5. What is the average speed based on age?

```{r}
res5 <- dbGetQuery(db,
                  "SELECT AVG( ((ABS(start_station_latitude-end_station_latitude) +
                                 ABS(start_station_longitude-end_station_longitude)) * 111) /
                            (TIMESTAMP_DIFF(stoptime, starttime, MINUTE)/60) ) AS avg_speed_km_h,
                          (2017 - birth_year) AS age
                  FROM citibike_trips
                  GROUP BY age
                  ORDER BY age")
```

```{r}
ggplot(data = subset(res5, age < 100), aes(x = age, y = avg_speed_km_h)) +
  geom_line() +
  labs(x = "Age", y = "Speed (km/h)",
       title = "Age vs. Average Speed (km/h)") +
  theme_minimal()
```

