---
title: "Querying an Google BigQuery database"
author: Pablo Barbera and Akitaka Matsuo
date: 27 November, 2018
output: html_document
---

You will now continue practicing how to write SQL queries using a Google BigQuery. There are a number of public databases available for testing ([link](https://bigquery.cloud.google.com/welcome/sharp-harbor-223622)). We will use two of such public databases.

## London LSOA Crime statistics

https://bigquery.cloud.google.com/table/bigquery-public-data:london_crime.crime_by_lsoa
This data counts the number of crimes at two different geographic levels of London (LSOA and borough) by year, according to crime type.


```{r}
library(bigrquery)
project <- "my472-bigquery"

get_table(project="bigquery-public-data",
          dataset="london_crime",
          table="crime_by_lsoa")
```
1. First, connect to the database and count how many rows it contains:
```{r}
query_exec("SELECT COUNT(*) FROM [bigquery-public-data:london_crime.crime_by_lsoa]",
           project = project, useLegacySql = TRUE)
```
2. Crime by year
```{r}
query_exec("SELECT year, SUM(value) AS count_crime FROM
           [bigquery-public-data:london_crime.crime_by_lsoa]
           GROUP BY year",
           project = project, useLegacySql = TRUE)
```

2. Crime by borough and year
```{r}
dat <- query_exec("SELECT year, borough, SUM(value) AS count_crime FROM
           [bigquery-public-data:london_crime.crime_by_lsoa]
           GROUP BY year, borough",
           project = project, useLegacySql = TRUE)
View(dat)
library(ggplot2)
library(tidyverse)
dat %>% group_by(borough) %>%
  mutate(ratio = count_crime/count_crime[year == 2008]) %>%
ggplot(aes(x = year, y = ratio, group = borough)) + geom_line()
```

3. Crime by year and category

```{r}
dat_3 <- query_exec("SELECT year, major_category, SUM(value) AS count_crime FROM
           [bigquery-public-data:london_crime.crime_by_lsoa]
           GROUP BY year, major_category",
           project = project, useLegacySql = TRUE)
View(dat_3)
ggplot(dat_3) + aes(x = year, y = count_crime, group = major_category,
                    colour = major_category) + geom_line()
```

4. Map?
```{r}
require(rgdal)
require(ggplot2)

dat_lsoa <- query_exec("SELECT lsoa_code, SUM(value) AS count_crime FROM
           [bigquery-public-data:london_crime.crime_by_lsoa]
           GROUP BY lsoa_code",
           project = project, useLegacySql = TRUE)
View(dat_lsoa)


# shapefile is from: https://data.london.gov.uk/dataset/statistical-gis-boundary-files-london
shp <- readOGR(dsn = "~/Downloads/statistical-gis-boundaries-london/ESRI/LSOA_2011_London_gen_MHW.shp", stringsAsFactors = F)
shp@data <- shp@data %>%
  mutate(id = rownames(shp@data)) %>%
  merge(dat_lsoa, by.x = "LSOA11CD", by.y = "lsoa_code")

shp_fort <- fortify(shp) %>%
  merge(shp@data, by = 'id')

map <- ggplot() + geom_polygon(data = shp_fort, aes(x = long, y = lat, group = group), colour = NA, fill = "red")
map

map <- ggplot() + geom_polygon(data = shp_fort, aes(x = long, y = lat, group = group,
                                               fill = count_crime), colour = NA)
map + theme_void()
```

## NYC Bicycle Hire

The second database is the one that contains all Citi Bike trips (NYC's bike-sharing service) since Citi Bike launched in September 2013. 

1. First, connect to the database and count how many rows it contains:

```{r}

get_table(project="bigquery-public-data",
          dataset="new_york",
          table="citibike_trips")
```

```{r}
# how many trips in database?
query_exec(
  "SELECT COUNT(*) FROM [bigquery-public-data:new_york.citibike_trips]",
	project = project, useLegacySql = TRUE)

# see first line
query_exec(
  "SELECT *
  FROM [bigquery-public-data:new_york.citibike_trips]
  LIMIT 10",
	project = project, useLegacySql = FALSE)
```

2. Which are the 10 most popular stations in terms of how many trips started there?

```{r}
q <- "SELECT start_station_id, COUNT(*) AS count_start
  FROM [bigquery-public-data:new_york.citibike_trips]
    GROUP BY start_station_id
  ORDER BY count_start DESC
  LIMIT 10"
query_exec(q, project = project, useLegacySql = FALSE)

```

3. What is the average trip duration based on the hour of the day when they start?

```{r}
query_exec(
  "SELECT HOUR(starttime) AS hour_of_day,
    AVG( (stoptime-starttime)/1000000/60 ) AS duration_minutes
  FROM [bigquery-public-data:new_york.citibike_trips]
  GROUP BY hour_of_day
  ORDER BY hour_of_day",
  project = project
)
```

4. What is the average trip duration of a CitiBike trip in NYC?

```{r}
query_exec(
  "SELECT AVG( (stoptime-starttime)/1000000/60 ) AS duration_minutes,
  FROM [bigquery-public-data:new_york.citibike_trips]",
  project = project
)
```

5. What is the average trip duration based on the age of the riders?

```{r}
res <- query_exec(
  "SELECT (2017-birth_year) AS age,
    AVG( (stoptime-starttime)/1000000/60 ) AS duration_minutes,
    COUNT(*) AS trip_count
  FROM [bigquery-public-data:new_york.citibike_trips]
  GROUP BY age
  ORDER BY age",
  project = project
)
plot(res$age[res$age<100], res$duration_minutes[res$age<100], type="l")
barplot(res$trip_count[!is.na(res$age) & res$age<100], 
        names.arg=res$age[!is.na(res$age) & res$age<100])
```

6. What is the average distance of a CitiBike trip in NYC?

```{r}
query_exec(
  "SELECT AVG( (ABS(start_station_latitude-end_station_latitude) +
      ABS(start_station_longitude-end_station_longitude)) * 111) AS avg_distance_km
  FROM [bigquery-public-data:new_york.citibike_trips]",
  project = project)
```

7. What is the average distance of a trip based on gender?

```{r}
query_exec(
  "SELECT AVG( (ABS(start_station_latitude-end_station_latitude) +
      ABS(start_station_longitude-end_station_longitude)) * 111) AS avg_distance_km,
    gender AS gender
  FROM [bigquery-public-data:new_york.citibike_trips]
  GROUP BY gender",
  project = project)
```

8. What is the average speed of a CitiBike trip?

```{r}
query_exec(
  "SELECT AVG( 
      ((ABS(start_station_latitude-end_station_latitude) + 
        ABS(start_station_longitude-end_station_longitude)) * 111) /
    ((stoptime-starttime)/1000000/3600) ) AS avg_speed_km_h
  FROM [bigquery-public-data:new_york.citibike_trips]",
  project = project)
```

9 What is the average speed based on age?

```{r}
res <- query_exec(
  "SELECT AVG( 
      ((ABS(start_station_latitude-end_station_latitude) + 
        ABS(start_station_longitude-end_station_longitude)) * 111) /
    ((stoptime-starttime)/1000000/3600) ) AS avg_speed_km_h,
    (2017-birth_year) AS age
  FROM [bigquery-public-data:new_york.citibike_trips]
  GROUP BY age
  ORDER BY age",
  project = project)

plot(res$age[res$age<100], res$avg_speed_km_h[res$age<100], 
     xlab="age", ylab="speed (km/h)", type="l")
```

