---
title: "MY472 - Week 2: Conditionals, loops, and functions"
author: "Ryan Hübert (AT 2024)"
date: "October 2024"
output: html_document
---

In this file, we will continue to review a few key concepts which you have seen 
in the preparatory course.

**Attribution statement:** _The following teaching materials have been 
iteratively developed by current and former instructors, including: Ken Benoit, 
Friedrich Geiecke, Thomas Robinson and Ryan Hübert._

### Conditionals

Conditionals execute blocks of code when certain logical conditions are met or 
not met. Recall the following list of logical operators also discussed in the 
first week:

```{r}
1 != 2 # not equal to
2 < 2 # less than
2 <= 2 # less than or equal to
2 > 2 # greater than
2 >= 2 # greater than or equal to
(2 < 2) | (2 <= 2) # or
(2 < 2) & (2 <= 2) # and
2 %in% c(0,3,24) # group membership
is.na(NA) # is NA
!is.na(42) # is not NA
```

Using these with "if" and "else" statements allows to run blocks of code 
depending on whether a logical condition is met or not, i.e. evaluates to true 
or false. In this course we might for example use conditionals while scraping to 
run different code depending on which site is currently accessed. The following 
simple example illustrates an if statement:

```{r}
x <- 2
if (x > 4) {
  print("If code block reached, condition was true.")
  exemplary_computation <- 73*14
  }
```


```{r}
x <- 5
if (x > 4) {
  print("If code block reached, condition was true.")
  exemplary_computation <- 73*14
  }
```

There might also be an alternative block of code we would like to run in case 
the condition is not met. This can be achieved with an if-else structure:

```{r}
x <- 3
if (x > 4) {
  print("If code block reached, condition was true.")
  exemplary_computation <- 73*14
} else {
  print("Else code block reached, condition was false.")
  exemplary_computation <- 37*41
  }
```

Lastly, there can be multiple if conditions and one else condition for all other cases:

```{r}
x <- 2
y <- 3
if (x < y) {
    print("First if code block reached, associated condition was true: y is greater than x.")
  } else if (x > y) {
    print("Second if code block reached, associated condition was true: x is greater than y.")
  } else {
    print("Else code block reached, none of the condition were true: x and y are equal.")
  }
```


### Loops

We use loops whenever we need to run the same function (or chunk of code) across 
different units. For example, in this course we may use a loop when we have 
multiple Twitter accounts and we want to run sentiment analysis for tweets 
posted by each of them, or when scraping data from multiple elements of a webpage.

"For" loops are probably the most common type of loop and are easily implemented in R:

```{r}
for (i in 1:10) {
    print(i)
}
```

Note the structure: 

```{r, eval = FALSE}
for (i in VECTOR) {
    # do something with i but not necessarily
}
```

In each iteration, i takes a different value of the VECTOR; "i" can be anything, 
we can e.g. also just name it number:

```{r}
for (number in 1:10) {
    print(number)
}
```

We can also iterate over different vectors than integers sequences with for 
loops:

```{r}
vector_of_texts <- c("hello", "world", "in", "a", "for", "loop")
for (text in vector_of_texts) {
    print(text)
}
```

Another type of loop is the while loop, it runs for as long as a logical 
condition is true:

```{r}
x <- 1
while (x < 11) {
    print(x)
    x <- x + 1
}
```

This while loop mimics the for loop above. More generally, while loops are more 
flexible. For example, you might run a while loop until a certain user input is 
given and sets a condition to false. One thing to be careful about with while 
loops is not to create conditions which are always true. This leads to infinite 
loops. For example "while (2 < 4) {print("hello word")}" would run forever.

A nice feature of loops is that we can use values from the  previous iteration. 
For instance, we can get the first 40 terms in the Fibonacci sequence using a 
for loop.

```{r}
fib <- c(0, 1, rep(NA, 38)) # initialize fib sequence with a vector (0, 1, NA, NA, NA, NA, ...)
for (i in 3:40) {
  fib[i] <- fib[i - 1] + fib[i - 2]
}
fib
```

Note that here we created an empty vector to store the output of each iteration. 
A simpler example:

```{r}
values <- rep(NA, 10)
for (i in 1:10) {
    values[i] <- i
}
values
```

A structure that we will use often in this course is a loop that stores some 
data in different elements within a list. This will be very useful when the 
output from each iteration is a data frame, for example while downloading data 
from an API. For example:

```{r}
# Create empty list
grades <- list()
# Loop over 5 students
for (i in 1:5) {
  # Create data frame with grade/info for this student
  student <- data.frame(id = i,
                        initial = sample(LETTERS, 1),
                        grade = runif(n = 1, min = 0, max = 100),
                        stringsAsFactors = FALSE)
  grades[[i]] <- student
}
# Now we have a list of single line data frames...
class(grades)
# We can combine these into a joint data frame
grades <- do.call(rbind, grades) # also see `bind_rows` from `dplyr` once we have discussed this package
grades
```

### Functions

Functions are objects just like the vectors and data frames we have worked with, 
so we create them using an assignment. The following function works for a single 
element as well as element-wise on a vector, because the * operator is defined 
element-wise.

```{r}
times_2 <- function(x) {x * 2}
times_2(6)
times_2(1:5)
```

We can also have multivariate inputs into functions, and return more complex 
objects, such as a vector or list.

```{r}
two_numbers <- function(x, y) {
  my_sum <- x + y
  my_product <- x * y
  my_ratio <- x / y
  return(c(my_sum, my_product, my_ratio))
}
two_numbers(4, 11.93)
```

A key conceptual feature of functions is that they have their own scope. What do 
you think the following cell will return?

```{r}
my_function <- function(z) {
  z <- 2*z
  return(z)
}
z <- 10

# First output
my_function(z)

# Second output
z
```

For more discussion on scoping, see e.g. 
<https://adv-r.hadley.nz/functions.html#lexical-scoping>

In modular code, particularly repeated tasks are written as functions. Functions 
therefore contain a wide range of possible code, e.g. conditions which allow the 
function to run different chunks depending on the input.

```{r}
compare_xy <- function(x, y) {
  if (x < y) {
    print("y is greater than x")
  } else if (x > y) {
    print("x is greater than y")
  } else {
    print("x and y are equal")
  }
}
```


```{r}
compare_xy(3, 4)
compare_xy(4, 3)
compare_xy(1, 1)
```

A slightly different type of if statement is the `ifelse` function:

```{r}

# Using an ifelse function to return an absolute value
x <- -2
abs_x <- ifelse(x > 0, x, -x)
abs_x

# This also works with a vector element-wise
numbers <- c(-2, -1, 0, 1, 2)
# converting them to absolute numbers
abs_numbers <- ifelse(numbers > 0, numbers, -numbers)
abs_numbers
```

A multi-condition version of `ifelse` which is nevertheless vectorised is the 
`case_when` function from the `dplyr` package, which is loaded as one of the 
packages in `tidyverse`:

```{r}
library(tidyverse)

# Using case_when to return fizz-buzz (a classic coding exercise)
numbers <- 1:15

# We can get the remainder of a division using %%
paste0("The remainder of 5/3 = ",5%%3)
paste0("The remainder of 5/5 = ",5%%5)

# case_when lets conditions be evaluated sequentially *until* one is met,
# for each element of a vector
fizzbuzz <- case_when(numbers %% 3 == 0 & numbers %% 5 == 0 ~ "fizzbuzz",
                      numbers %% 3 == 0 ~ "fizz",
                      numbers %% 5 == 0 ~ "buzz",
                      TRUE ~ "") # note, it's good practice to have a default
print(fizzbuzz)
```

Further study: A great book to review fundamental and advanced concepts in R 
programming is Hadley Wickham's "Advanced R" <https://adv-r.hadley.nz/>