---
title: "Week 2: Seminar exercise"
author: "Martin Lukac"
date: "10/8/2020"
output: html_document
---

You will use the `tidyverse` package and the dataset called `ip_and_unemployment.csv` that we used in lectures. Start with setting up your workspace:

```{r setup}
library(tidyverse)
df = read.csv('ip_and_unemployment.csv', fileEncoding="UTF-8-BOM")
df_wide = pivot_wider(df, names_from = series, values_from = value)
```
    
What are the highest unemployment rates for France and Spain during the time of the sample?  What are the lowest values for monthly percentage changes of industrial production for the two countries? Make sure to delete NA values in only the time series of interest. (_Optional_: can you create a function that would do this for any country?)

```{r q1}
# Q1 --------------------------------------------------------------------------

# France

df_wide %>% 
  filter(country == 'france') %>%
  filter(unemployment == max(unemployment, na.rm = TRUE) | ip == min(ip, na.rm = TRUE))

# Spain 

df_wide %>% 
  filter(country == 'spain') %>%
  filter(unemployment == max(unemployment, na.rm = TRUE) | ip == min(ip, na.rm = TRUE))


# Optional - general function

filter_worst_month <- function( x ) {
  filtered_df <- df_wide %>%
    filter(country == x) %>%
    filter(unemployment == max(unemployment, na.rm= TRUE) | ip== min(ip, na.rm=TRUE))
  return(filtered_df)
}

filter_worst_month('germany')
```


-----

How many non-NA monthly observations of industrial production exist for the countries here. Can you  determine this with the summarise and group_by functions? (_Optional_: can you calculate the % of values that are non-NA?)


```{r q2} 
# Q2 --------------------------------------------------------------------------

monthly_df <- df_wide %>%
  group_by(country) %>%
  select(country, ip) %>%
  drop_na() %>%
  summarise(observations_non_na = n())
monthly_df
```
```{r q2.0}
df_wide %>% 
  group_by(country) %>%
  select(country,ip) %>%
  summarise(means=mean(ip, na.rm=TRUE), observations = n())

# Experimenting alone: 
# I check the number of NA-s in each group, to approach the question from this angle.
monthly_na <- df_wide %>%
  group_by(country) %>%
  select(country, ip) %>%
  summarise(observations_all = n(), num_na = sum(is.na(ip)))
monthly_na

correct = identical(monthly_na$observations_all - monthly_na$num_na, monthly_df$observations_non_na)
correct
# Indeed, subtracting the number of NA-s from the total number of observations yields the same as counting non-NAs directly.


# Martin's solution

ipu_clean <- df_wide %>%
  group_by(country) %>%
  summarise(nonNA_ip = sum(!is.na(ip)),
            nonNA_ue = sum(!is.na(unemployment)))
ipu_clean

ipu_clean_2 <- df_wide %>%
  group_by(country) %>%
  summarise(nonNA_ip = sum(!is.na(ip)),
            nonNA_ue = sum(!is.na(unemployment)),
            nonNA_ip_pct = nonNA_ip / length(ip),
            nonNA_ue_pct = nonNA_ue / length(unemployment))
ipu_clean_2

```

```
-----

In data science and machine learning, it can sometimes increase the predictive power of models to add transformations of existing variables. This is usually done in the modelling step, but to practice using the `mutate` function, let's do it here. Add three new columns to the dataframe: 

  1. the square of the industrial production percentage change, 
  2. the natural logarithm of the unemployment rate, and 
  3. the interaction (i.e. the product) of industrial production percentage change and unemployment rate.
  
(_Optional_: Calculate the difference between last year's and this year's unemployment rate and industrial production. Are we faring worse or better than last year?)

```{r q3}
# Q3 --------------------------------------------------------------------------

df_wide_extended <- df_wide %>%
  mutate(ip_percentage_change_squared = (ip/lag(ip)-1)*100) %>%
  mutate(ln_unemployment = log(unemployment)) %>%
  mutate(interaction_of_ip_and_unemployment = (ip/lag(ip)-1)*100 * unemployment)
  
df_wide_extended_2 <- df_wide %>%
  mutate(ip_sq = (ip/lag(ip)-1)*100,
          unemployment_ln = log(unemployment),
          ip_unemployment = ip * unemployment) %>%
  head()

df_wide_extended
df_wide_extended_2
# Is there a way to do this without using mutate only once?


# Optional
# I will calculate monthly comparisons as a first approach

df_uk <- df_wide %>%
  filter(country == 'uk') %>%
  mutate(diff_unemp = unemployment - lag(unemployment, n=12L), na.rm=TRUE) %>%
  mutate(diff_ip = ip - lag(ip, n=12L), na.rm=TRUE) %>%
  select(date, diff_unemp, diff_ip)%>%
  filter(!is.na(diff_ip)|!is.na(diff_unemp))
 
df_uk

#dates <- as.Date(df_uk$date, "%d.%m.%Y")
#year_month <- format(dates, format= '%Y.%m.')
#year_month


```
