---
title: "Reshaping datasets"
author: "Kenneth Benoit and Martin Lukac"
date: "05/10/2020"
output:
  html_document: default
  pdf_document: default
---

We often want to convert the format in which a data frame is structured, from wide to long format, or vice versa. Let's see some examples of how to do this:

```{r}
d <- read.csv("unvoting.csv", stringsAsFactors = FALSE)
dd <- d[d$CountryAbb %in% c("HUN", "AUS") & d$Year %in% 2010:2012,
        c("Year", "CountryAbb", "idealpoint")]

# converting to wide format
library(reshape2)
(dc <- dcast(dd, CountryAbb ~ Year, value.var = "idealpoint"))

# converting back to long format
melt(dc, id.vars = c("CountryAbb"), value.name = "idealpoint")
```

Another very useful function to reshape datasets (although it's more about computing summary statistics!) is `aggregate`. For example, we can use it to compute the average value of a variable, grouped by values of a different variable. Let's see some examples

```{r}
aggregate(idealpoint ~ CountryAbb, data = d, FUN = mean)

aggregate(idealpoint ~ CountryAbb, data = d, FUN = median)

aggregate(PctAgreeRUSSIA ~ Year, data = d, FUN = mean)
plot(aggregate(PctAgreeRUSSIA ~ Year, data = d, FUN = mean), type = "l")

aggregate(idealpoint ~ CountryAbb, data = d, FUN = length)

```

## Merging datasets

Two or more data frames can be combined into a single one according to the values of a single variables using the `merge` function.

```{r}
students <- data.frame(name = c("Paul", "Peter", "Amy", "Megan", "Selena"),
                age = c(18, 19, 18, 20, 21),
                stringsAsFactors = FALSE)
grades <- data.frame(name = c("Amy", "Megan", "Peter", "Drew"),
                     grade = c("A", "B+", "A-", "B"),
                     stringsAsFactors = FALSE)

merge(students, grades)
# we can be more specific about how to merge
merge(students, grades, by = "name")

# if the common variable has different names, we need to specify it
grades <- data.frame(student = c("Amy", "Megan", "Peter", "Drew"),
                     grade = c("A", "B+", "A-", "B"),
                     stringsAsFactors = FALSE)
merge(students, grades, by.x = "name", by.y = "student")

```

Note that by default `merge` will only return rows that are present in both dataset; but we can change that default behavior. For the non-matches, the returned value will be `NA`.

```{r}
merge(students, grades, by.x = "name", by.y = "student", all.x = TRUE)
merge(students, grades, by.x = "name", by.y = "student", all.y = TRUE)
merge(students, grades, by.x = "name", by.y = "student", all = TRUE)
```

What happens if names do not have exactly the same spelling? We can try some fuzzy string matching based on the Levenshtein (or edit) distance between strings. See `?adist`

```{r}
students <- data.frame(name = c("Paul", "Pete", "Amie", "Meghan", "Selena"),
                age = c(18, 19, 18, 20, 21),
                stringsAsFactors = FALSE)
grades <- data.frame(name = c("Amy", "Megan", "Peter", "Drew"),
                     grade = c("A", "B+", "A-", "B"),
                     stringsAsFactors = FALSE)

# probably there are better ways to code this...
students$grade <- NA

for (i in seq_len(nrow(students))) {
  dist <- adist(students$name[i], grades$name)
  if (any(dist < 4)) {
    students$grade[i] <- grades$grade[which.min(dist)]
  }
}

students

```

How does it work? What can you do if you have a loop that you don't understand and need to grasp what it does at each step?

Try printing messages with `message()` or `print()`. It is not beautiful, but it is very efficient way of debugging and analysing your code.

For example, to understand the performance of the last loop:

```{r}
students <- data.frame(name = c("Paul", "Pete", "Amie", "Meghan", "Selena"),
                age = c(18, 19, 18, 20, 21),
                stringsAsFactors = FALSE)
grades <- data.frame(name = c("Amy", "Megan", "Peter", "Drew"),
                     grade = c("A", "B+", "A-", "B"),
                     stringsAsFactors = FALSE)
students$grade <- NA

# added messages in the loop to show what it does at each step
for (i in seq_len(nrow(students))) {
  dist <- adist(students$name[i], grades$name)
  message(i, ". matching ", students$name[i], ":\n",              # what student
          "    distances from others: ", dist[1,], "\n")          # distances
  if (any(dist < 4)) {
    students$grade[i] <- grades$grade[which.min(dist)]
    message("    matching with ", grades$name[which.min(dist)])   # outcome A
  } else {
    message("    ", students$name[i], " is unmatched!")           # outcome B
  }
}
```


To learn more about merging datasets based on strings, see the `RecordLinkage` package.

Another package that is very useful is `countrycode`, which contains a data frame with pretty much every potential country code you may want to use.
