---
title: "Querying large databases via Google BigQuery"
author: "Pablo Barbera, Friedrich Geiecke, Akitaka Matsu"
date: "30 November 2020"
output: html_document
---

In this file we will continue practicing how to query online databases with SQL. We will use our BigQuery setup from the previous document with two public databases to demonstrate queries for large datasets.

Loading packages:

```{r}
library("DBI")
library("bigrquery")
library("tidyverse")
library("rgdal")
```

## London LSOA Crime statistics

This dataset contains the number of crimes at two different geographic levels of London (LSOA and borough) by year, according to crime type. First, we have to change our project name such that the `bigrquery` package knows that it has to request the public datasets now.

```{r}
# Project name (now accessing the public datasets)
project_name <- "bigquery-public-data"

# Billing (main project ID)
billing_info <- "tba"

# Dataset name (London crime database within the public databases)
dataset_name <- "london_crime"
```

Creating the database object:

```{r}
db <- dbConnect(
  bigrquery::bigquery(),
  project = project_name,
  dataset = dataset_name,
  billing = billing_info
)
db
```

1. First, connect to the database and count how many rows it contains (if the code yields a Google credentials error, once copy-paste it into the R console directly which should then open the browser window for authentication):

```{r}
dbGetQuery(db, "SELECT COUNT(*) FROM crime_by_lsoa")
```

2. Crime by year

```{r}
dbGetQuery(db, "SELECT year, SUM(value) AS count_crime FROM crime_by_lsoa
           GROUP BY year
           ORDER BY year")
```

3. Crime evolution by borough and year

```{r}
data_plot_1 <- dbGetQuery(db, "SELECT year, borough, SUM(value) AS count_crime
            FROM crime_by_lsoa
            GROUP BY year, borough")
View(data_plot_1)
```

```{r fig.width=7, fig.height=4, echo=FALSE}
data_plot_1 %>% group_by(borough) %>%
  mutate(ratio = count_crime/count_crime[year == 2008]) %>%
ggplot(aes(x = year, y = ratio, color = borough)) + geom_line()
```

4. Crime by year and category

```{r}
data_plot_2 <- dbGetQuery(db, "SELECT year, major_category,
            SUM(value) AS count_crime FROM crime_by_lsoa
            GROUP BY year, major_category")
#View(data_plot_2)
```

```{r}
ggplot(data_plot_2) + aes(x = year, y = count_crime, colour = major_category) +
  geom_line()
```

5. Depicting crime on a map of the city:

```{r}
crime_per_lsoa_code <- dbGetQuery(db, "SELECT lsoa_code, SUM(value) AS count_crime
            FROM crime_by_lsoa
            GROUP BY lsoa_code")
crime_per_lsoa_code
```

```{r}

# Download the shape file from: https://data.london.gov.uk/dataset/statistical-gis-boundary-files-london
shp <- readOGR(dsn = "~/Downloads/statistical-gis-boundaries-london/ESRI/LSOA_2011_London_gen_MHW.shp",
               stringsAsFactors = FALSE)

# Obtaining longitude and latitude (contains an ID column, but starting at zero!)
shp_fort <- fortify(shp)

# Merging region information (such as names) contained in shp@data with
# a) this geographical information and
# b) the crime information

data_plot_3 <- shp@data %>%
  mutate(id = 0:(nrow(shp@data)-1)) %>% # add an id column also to shp@data
  merge(shp_fort, by = "id") %>% # merge with longitude and latitude data based on this column
  merge(crime_per_lsoa_code, by.x = "LSOA11CD", by.y = "lsoa_code") # merge with crime data

```

```{r}
map <- ggplot() +
  geom_polygon(data = data_plot_3, aes(x = long, y = lat, group = group),
               colour = NA, fill = "red")
map

map <- ggplot() +
  geom_polygon(data = data_plot_3, aes(x = long, y = lat, group = group,
                                    fill = count_crime), colour = NA)
map + theme_void()
```

## NYC Bicycle Hire

The second database contains Citi Bike trips (NYC's bike-sharing service) since Citi Bike launched in September 2013.

First, we need to update the dataset name and connect again:

```{r}
dataset_name <- "new_york"
```

```{r}
db <- dbConnect(
  bigrquery::bigquery(),
  project = project_name,
  dataset = dataset_name,
  billing = billing_info
)
db
```

1. First, connect to the database and count how many rows it contains:

```{r}
# Number of trips in the database
dbGetQuery(db, "SELECT COUNT(*) FROM citibike_trips")
```

```{r}
# First 10 entries
dbGetQuery(db, "SELECT * FROM citibike_trips LIMIT 10")
```

2. Which are the 10 most popular stations in terms of how many trips started there?

```{r}
dbGetQuery(db, "SELECT start_station_name, COUNT(*) AS count_start
  FROM citibike_trips
  GROUP BY start_station_name
  ORDER BY count_start DESC
  LIMIT 10")
```

3. What is the average trip duration of a CitiBike trip in NYC? For advanced keywords like these, see the detailed [documentation](https://cloud.google.com/bigquery/docs/reference/standard-sql/timestamp_functions).

```{r}
dbGetQuery(db, "SELECT AVG( TIMESTAMP_DIFF(stoptime, starttime, MINUTE) )
           AS duration_minutes
           FROM citibike_trips")
```

4. What is the average trip duration based on the hour of the day when they start?

```{r}
dbGetQuery(db, "SELECT EXTRACT(HOUR FROM starttime) as hour_of_day,
  AVG( TIMESTAMP_DIFF(stoptime, starttime, MINUTE) ) AS duration_minutes
  FROM citibike_trips
  GROUP BY hour_of_day
  ORDER BY hour_of_day")
```

5. What is the average trip duration based on the gender of the riders?

```{r}
dbGetQuery(db, "SELECT gender, AVG( TIMESTAMP_DIFF(stoptime, starttime, MINUTE) )
  AS duration_minutes, COUNT(*) as trip_count,
  FROM citibike_trips
  GROUP BY gender")
```

6. What is the average distance of a trip?

```{r}
dbGetQuery(db, "SELECT AVG( (ABS(start_station_latitude-end_station_latitude) +
      ABS(start_station_longitude-end_station_longitude)) * 111) AS avg_distance_km
  FROM citibike_trips")
```

7. What is the average distance of a trip based on gender?

```{r}
dbGetQuery(db, "SELECT gender, AVG( (ABS(start_station_latitude-end_station_latitude) +
      ABS(start_station_longitude-end_station_longitude)) * 111) AS avg_distance_km
  FROM citibike_trips
  GROUP BY gender")
```





















