---
title: "Week 10: Further ways to query an SQLite database"
date: "Autumn Term 2024"
output: html_document
---

**Attribution statement:** _The following teaching materials have been iteratively developed by current and former instructors, including: Pablo Barberá and Ryan Hübert._

```{r}
library(DBI)
```

First, let's do a little directory management. Where do we want to store our work for this lecture? Indicate the path below, naming it `ddir`. Then, be sure to put the folder `facebook-data` inside this folder.

```{r}
# Where is your data stored on your computer for this week's lecture?
ddir <- "tba" 
```

## Joining and querying multiple tables

Now is where we get to see the magic of SQL. It makes it really easy to combine multiple SQL tables using **`JOIN`**. We just select columns from different tables and then specify what _key_ should be used to match them using **`ON`**. 

```{r}
db <- dbConnect(RSQLite::SQLite(), paste0(ddir, "facebook-db.sqlite"))

dbGetQuery(db, 
  "SELECT posts.likes_count, congress.party, posts.date
  FROM posts JOIN congress
    ON congress.screen_name = posts.screen_name
  LIMIT 10")
```

Once we know how to merge tables, we can use columns from both tables. For example, let us say we want to find the most shared post by a Democrat, or by a Republican (more on this in detail in this file). Here is how we would do it:

```{r}
dbGetQuery(db, 
  "SELECT posts.from_name, posts.message, posts.shares_count, congress.party 
  FROM posts JOIN congress
    ON congress.screen_name = posts.screen_name
  WHERE party = 'Democrat'
  ORDER BY shares_count DESC
  LIMIT 1")
dbGetQuery(db, 
  "SELECT posts.from_name, posts.message, posts.shares_count, congress.party 
  FROM posts JOIN congress
    ON congress.screen_name = posts.screen_name
  WHERE party = 'Republican'
  ORDER BY shares_count DESC
  LIMIT 1")
```

Note that `JOIN` per default performs an `INNER JOIN`, however, it can be clearer to specify the type of join as there are several (left vs right outer joins, ...). We will discuss joins in more detail in this week's class.

## Grouping and aggregating

SQL makes it very easy to compute aggregate statistics with a series of built-in functions (COUNT, AVG, MAX, MIN...). Let's now see a few examples of this type of queries:

```{r}
# Use COUNT with * to return the total number of rows in a table
dbGetQuery(db, "SELECT COUNT(*) FROM posts")
dbGetQuery(db, "SELECT COUNT(*) FROM congress")
```

A key function for aggregation is **`GROUP BY`** which allows to aggregate results based on values of a specific variables.

```{r}
dbGetQuery(db, 
  "SELECT from_name, COUNT(*) AS post_count
  FROM posts
  GROUP BY from_name
  LIMIT 10")
```

We can then e..g. order the results using **`ORDER BY`**. The default is alphabetical order for the first column, but we can change that:

```{r}
# Sort by type_count
dbGetQuery(db, 
  "SELECT type, COUNT(type) AS type_count
  FROM posts
  GROUP BY type
  ORDER BY type_count")

# In descending orders
dbGetQuery(db, 
  "SELECT type, COUNT(type) AS type_count
  FROM posts
  GROUP BY type
  ORDER BY type_count DESC")

# Which was the most popular post?
dbGetQuery(db, 
  "SELECT from_name, message, likes_count, datetime
  FROM posts
  ORDER BY likes_count DESC
  LIMIT 1")

# You can also specify the column number instead of the name
dbGetQuery(db, 
  "SELECT from_name, message, likes_count, datetime
  FROM posts
  ORDER BY 3 DESC
  LIMIT 1")

# What was the post with the highest comment to like ratio?
# We subset only posts with 1000 likes or more
dbGetQuery(db,
  "SELECT from_name, message, likes_count, comments_count, date,   
      comments_count/likes_count AS comment_like_ratio
  FROM posts
  WHERE likes_count > 1000
  ORDER BY comment_like_ratio DESC
  LIMIT 5")
```

While we can already obtain a lot of information about posts, we cannot yet see e.g. the party as it is contained in the congress table rather than posts table. Using join, we can for example count the total number of posts by party, or by party and gender.

```{r}
dbGetQuery(db, 
  "SELECT congress.party, COUNT(*) AS total_posts
  FROM posts JOIN congress
    ON congress.screen_name = posts.screen_name
  GROUP BY congress.party")

dbGetQuery(db, 
  "SELECT congress.party, congress.gender, COUNT(*) AS total_posts
  FROM posts JOIN congress
    ON congress.screen_name = posts.screen_name
  GROUP BY congress.party, congress.gender")

```

Here is another example. Let us say we want to look at what domains Republicans and Democrats share more frequently on their Facebook pages. How would we do this?

```{r}
dbGetQuery(db, 
  "SELECT congress.party, domain, COUNT(*) AS domain_count
  FROM posts JOIN congress
    ON congress.screen_name = posts.screen_name
  WHERE congress.party = 'Democrat'
  GROUP BY domain
  ORDER BY domain_count DESC
  LIMIT 20")

dbGetQuery(db, 
  "SELECT congress.party, domain, COUNT(*) AS domain_count
  FROM posts JOIN congress
    ON congress.screen_name = posts.screen_name
  WHERE congress.party = 'Republican'
  GROUP BY domain
  ORDER BY domain_count DESC
  LIMIT 20")

```

There are other functions besides just **`COUNT`**. We can also compute averages, the maximum and minimum of a column, the sum... We can also use functions like **`SUBSTR`** to extract parts of a character string.

```{r}
# Average number of posts by party
dbGetQuery(db, 
  "SELECT congress.party, AVG(posts.likes_count), COUNT(*) 
  FROM posts JOIN congress
    ON congress.screen_name = posts.screen_name
  GROUP BY congress.party")

# We can also compute it in a different way...
dbGetQuery(db, 
  "SELECT congress.party, SUM(posts.likes_count)/COUNT(*) AS average 
  FROM posts JOIN congress
    ON congress.screen_name = posts.screen_name
  GROUP BY congress.party")

# Most popular post by party
dbGetQuery(db, 
  "SELECT from_name, message, congress.party, MAX(posts.likes_count), COUNT(*) 
  FROM posts JOIN congress
    ON congress.screen_name = posts.screen_name
  GROUP BY congress.party")

# Number of posts by day of the month (recall, the date format is: YYYY-MM-DD)
dbGetQuery(db, 
  "SELECT SUBSTR(date, 9, 2) AS day_of_month, COUNT(*) as post_count
  FROM posts
  GROUP BY day_of_month")

```


```{r}
# Average number of likes and comments, per candidate using a CTE with aliasing (just for demonstration)
dbGetQuery(db, 
  "WITH 
    
    fb_metrics AS(
      SELECT screen_name, 
      AVG(likes_count) AS avg_likes_count, 
      AVG(comments_count) AS avg_comments_count
    FROM posts
    GROUP BY screen_name  
    )
  
  SELECT
    c.screen_name,
    fm.avg_likes_count,
    fm.avg_comments_count
  FROM congress c
  INNER JOIN fb_metrics fm
  ON c.screen_name = fm.screen_name;
  ")

```