---
title: "MY472 - Week 10: Seminar exercises on BigQuery"
author: "Pablo Barberá, Friedrich Geiecke"
date: "30 November 2023"
output: html_document
---

## Data

We will use the NYC Bicycle Hire data we looked at in lecture. The database contains Citi Bike trips (NYC's bike-sharing service) since Citi Bike launched in September 2013.

```{r}
library("DBI")
library("bigrquery")
```

## Queries

1. First, connect to the database and see the first 5 entries.

```{r}
# Project name (now accessing the public datasets)
project_name <- "bigquery-public-data"

# Billing (main project ID)
billing_info <- "tba"

# Dataset name (part of the public datasets)
dataset_name <- _________
```

```{r}
db <- dbConnect(
  bigrquery::bigquery(),
  project = project_name,
  dataset = dataset_name,
  billing = billing_info
)
db
```

All tables in the database:

```{r}
dbListTables(db)
```

```{r}
# First 5 entries
dbGetQuery(db,
           "...")
```

2. What is the average trip duration based on the age of the riders?

```{r}

```

3. What is the average distance of a CitiBike trip based on the age of the riders?

```{r}

```

4. What is the average speed of a CitiBike trip?

```{r}

```

5. What is the average speed based on age?

```{r}

```

