# Using BigQuery

## Data

We will use the NYC Bicycle Hire data we looked at in lecture. The database contains all Citi Bike trips (NYC's bike-sharing service) since Citi Bike launched in September 2013. 


## Queries

1. First, connect to the database and see the first 5 entries.

```{r}
library(bigrquery)
project <- "mblukac-my472"

query_exec(
  "SELECT *
  FROM [bigquery-public-data:new_york.citibike_trips]
  LIMIT 5",
	project = project)
```

2. What is the average trip duration based on the age of the riders?

```{r}
res <- query_exec(
  "SELECT (2020-birth_year) AS age,
    AVG( (stoptime-starttime)/1000000/60 ) AS duration_minutes,
    COUNT(*) AS trip_count
  FROM [bigquery-public-data:new_york.citibike_trips]
  GROUP BY age
  ORDER BY age",
  project = project
)
plot(res$age[res$age<100], res$duration_minutes[res$age<100], type="l")
barplot(res$trip_count[!is.na(res$age) & res$age<100], 
        names.arg=res$age[!is.na(res$age) & res$age<100])
```

3. What is the average distance of a CitiBike trip based on the age of the riders?

```{r}
res <- query_exec(
  "SELECT (2017-birth_year) AS age,
      AVG( (ABS(start_station_latitude-end_station_latitude) +
      ABS(start_station_longitude-end_station_longitude)) * 111) AS avg_distance_km
  FROM [bigquery-public-data:new_york.citibike_trips]
  GROUP BY age
  ORDER BY age",
  project = project)
plot(res$age[res$age<100], res$avg_distance_km[res$age<100], type="l")
```

4. What is the average speed of a CitiBike trip?

```{r}
query_exec(
  "SELECT AVG( 
      ((ABS(start_station_latitude-end_station_latitude) + 
        ABS(start_station_longitude-end_station_longitude)) * 111) /
    ((stoptime-starttime)/1000000/3600) ) AS avg_speed_km_h
  FROM [bigquery-public-data:new_york.citibike_trips]",
  project = project)

```

5. What is the average speed based on age?

```{r}
res <- query_exec(
  "SELECT AVG( 
      ((ABS(start_station_latitude-end_station_latitude) + 
        ABS(start_station_longitude-end_station_longitude)) * 111) /
    ((stoptime-starttime)/1000000/3600) ) AS avg_speed_km_h,
    (2017 - birth_year) AS age
  FROM [bigquery-public-data:new_york.citibike_trips]
  GROUP BY age
  ORDER BY age",
  project = project)

plot(res$age[res$age<100], res$avg_speed_km_h[res$age<100], 
     xlab="age", ylab="speed (km/h)", type="l")
```



