---
title: "MY472 - Week 10: Seminar exercises on SQL subqueries"
author: "Pablo Barberá"
date: "30 November 2023"
output: html_document
---

## Facebook database

Use the SQLite database with Facebook posts data we created and used in week 9 to complete the questions below.

```{r}
library(DBI)
library(tidyverse)
library(microbenchmark)
```

## Subqueries

In SQL, you can run a nested query where you use a result from the first query as a condition for the second query. Here, we will find the top 3 most frequent posters in congress and then get all messages from them. 

```{r}
db <- dbConnect(RSQLite::_________, "data/facebook-db.sqlite")
```

```{r}
# Congress table
dbListFields(db, "congress")

# Posts table
dbListFields(db, "posts")
```

1. What are the screen names of the top 3 posters?

```{r}
dbGetQuery(db, 
           "...")
```

2. (Without subquery) how can you get the posts from these congress members?

```{r}

```

3. Now, instead of hardcoding the screen_names in the conditional statement, use the query results in #1 as the subquery. 

```{r}

```

4. Using a subquery, get the average number of likes per post for all women Democrats.

```{r}

```

5. Now use a join to do the same as in #4.

```{r}

```

6. Which solution is faster: the one in #4 or #5 (try to use `microbenchmark`)?

```{r}

```
