---
title: "Scraping table data"
author: "Pablo Barbera and Akitaka Matsuo"
date: October 25, 2018
output: html_document
---

### Scraping web data in table format

We will start by loading the `rvest` package, which will help us scrape data from the web.

```{r, message=FALSE}
library(rvest)
```

The goal of this exercise is to scrape the number of new social security number holders by year in the US, and then clean it so that we can generate a plot showing the evolution in this variable over time.

The first step is to read the html code from the website we want to scrape, using the `read_html()` function. If we want to see the html in text format, we can then use `html_text()`.

```{r}
url <- "https://www.ssa.gov/oact/babynames/numberUSbirths.html"
# reading the html code into memory

# not very informative

# first 1000 characters

```

To extract all the tables in the html code automatically, we use `html_table()`. Note that it returns a list of data frames, so in order to work with this dataset, we will have to subset the second element of this list.

```{r, eval = FALSE}

# check the tables

# use the population tables
pop <- 
```

Now let's clean the data so that we can use it for our analysis. We need to convert the population values into a numeric format, which requires deleting the commas. We will also change the variable names so that it's easier to work with them.

```{r, eval = FALSE}
## remove commas and then make variable numeric
pop$Male <-  
pop$Female <- 

## rename variables
names(pop) <- 
```

And now we can plot to see how the number of people applying for a Social Security Number in the US has increased over time.

```{r, eval = FALSE}
plot()
lines()
legend()

## or ggplot

```

### Scraping web data in table format: a more advanced example

When there are multiple tables on the website, scraping them becomes a bit more complicated. Let's work through a common case scenario: scraping a table from Wikipedia with a list of the most populated cities in the United States.

```{r, eval = FALSE}
url <- 'https://en.wikipedia.org/wiki/List_of_United_States_cities_by_population'
html <- read_html(url)
tables <- 
length(tables)
```

The function now returns 12 different tables. I had to use the option `fill=TRUE` because some of the tables appear to have incomplete rows.

In this case, identifying the part of the html code that contains the table is a better approach. To do so, let's take a look at the source code of the website. In Google Chrome, go to _View_ > _Developer_ > _View Source_. All browsers should have similar options to view the source code of a website.

In the source code, search for the text of the page (e.g. _2017 rank_). Right above it you will see: `<table class="wikitable sortable"...">`. This is the CSS selector that contains the table. (You can also find this information by right-clicking anywhere on the table, and choosing _Inspect_).

Now that we now what we're looking for, let's use `html_nodes()` to identify all the elements of the page that have that CSS class. (Note that we need to use a dot before the name of the class to indicate it's CSS.)

```{r, eval = FALSE}
wiki <- html_nodes()
length(wiki)
```

There are 7 tables in total, and we will extract the first one.
```{r, eval = FALSE}
pop <- 

```

As in the previous case, we still need to clean the data before we can use it. For this particular example, let's see if this dataset provides evidence in support of [Zipf's law for population ranks](https://en.wikipedia.org/wiki/Zipf%27s_law).

We'll use regular expressions to remove endnotes and commas in the population numbers, and clean the variable names. (We'll come back to this later in the course.)

```{r, eval = FALSE}
pop$city_name <- 
pop$population <- 
pop$population <- 
pop$rank <- 
pop <- pop[,c("rank", "population", "city_name")]
```

Now we're ready to generate the figure:

```{r, eval = FALSE}
library(ggplot2)
p <- ggplot(pop, aes(x=rank, y=population, label=city_name))
pq <- p + geom_point() + ...
```

