---
title: "ggplot2 Basics: Grammar of graphics, geom types"
date: "16/11/2020"
output: github_document
---

## Introduction

In the examples below, we will be working with a random sample of 10,000 public Facebook posts by members of the U.S. Congress. 

For a review of the log scale, see for example this [video](https://www.youtube.com/watch?v=sBhEi4L91Sg).

Loading packages:

```{r}
library("tidyverse")
library("lubridate")
library("scales")
```

Reading in the data and some initial processing:

```{r}
df <- read.csv("data/fb-congress-data.csv", stringsAsFactors = FALSE)

# Transform date column to datetime
df$date <- as_date(df$date)

# Dropping some very uncommon post types
nrow(df)
df <- df %>%
  filter(!post_type %in% c("music", "note"))
nrow(df)
```


## Time series

After creating a base layer, `geom_line()` can be used for line plots such as time series. Plotting daily posts over time:

```{r}
counts <- df %>%
  group_by(date) %>% 
  summarise(posts = n())

# Base layer
p <- ggplot(counts, aes(x = date, y = posts))

# Line plot of the posts per day
p + geom_line()
```


Two separate time series by party, now also aggregated to monthly:

```{r}
# Obtain a new data frame with monthly counts of posts per party
counts <- df %>%
  filter(party != "Independent") %>% 
  group_by(month = ceiling_date(date, "month"), party) %>% 
  summarise(posts = n())

p <- ggplot(counts, aes(x = month, 
                        y = posts, group = party))
p + geom_line(aes(color = party)) +
  scale_color_manual(values=c("blue", "red"))
```


## Univariate analysis for a single continuous variable

```{r}
# Base layer
p <- ggplot(df, aes(x=likes_count))

# Histogram
p + geom_histogram()

# Smoothed density estimate
p + geom_density() + scale_x_continuous(labels = comma)

# The same log scale (note: the labels = comma option prevents scientific
# notation of numbers as 1e+00, 1e+01 and uses the `scales` package)
p + geom_histogram() + scale_x_log10(labels = comma)
p + geom_density() + scale_x_log10(labels = comma)

# Why does this line of code drops some observations?
```

## Univariate analysis for a single categorical variable

```{r}
p <- ggplot(df, aes(x=post_type))

# Bar chart
p + geom_bar() ## number of posts by type

# Bar chart (horizontal)
p + geom_bar() + coord_flip()
```


## Bivariate analysis for two continuous variables

```{r}
# Base layer
p <- ggplot(df, aes(x = likes_count, y = comments_count))

# Scatter plot: Relationship between number of likes and number of comments
p + geom_point()  

# With smoothed conditional means
p + geom_point() + stat_smooth(na.rm = TRUE) 

# With restricted axes
p + geom_point() + xlim(0, 25000) + ylim(0, 2500)

# Particularities of integer variables in scatter plots
p + geom_point() + xlim(0, 10) + ylim(0, 10)
  
# With log scales
p + geom_point() + scale_x_log10(labels = comma) + scale_y_log10(labels = comma) 
p + geom_point() + scale_x_log10(labels = comma) + scale_y_log10(labels = comma) + 
    stat_smooth()
```


## Bivariate analysis for one continuous variable and one categorical variable

```{r}
# Number of likes by type of post as a box blot
p <- ggplot(df, aes(x = post_type, y = likes_count))
p + geom_boxplot()

# Box plot and violin plot with log scale
p + geom_boxplot() + scale_y_log10(labels = comma)
p + geom_violin() + scale_y_log10(labels = comma)

# Density plot for log like distributions for different parties
p <- ggplot(df, aes(x = likes_count))
p + geom_density(aes(color = party)) + scale_x_log10(labels = comma)
```


## Bivariate analysis for two categorical variables

```{r}
counts <- df %>%
  filter(party != "Independent") %>% 
  group_by(post_type, party) %>% 
  summarise(posts = n())

p <- ggplot(counts, aes(x = party, y = post_type))
p + geom_tile(aes(fill = posts))

```


## Multivariate analysis for three continuous variables

```{r}
p <- ggplot(df, aes(x = likes_count, y = comments_count, color = log(angry_count)))

p + geom_point()

p + geom_point() + scale_y_log10(labels = comma) + scale_x_log10(labels = comma) + 
 stat_smooth(method = "lm")
```


## Multivariate analysis for two continuous variables and one categorical variable

```{r}

# Grid of plots: 2x4, by post type
p <- ggplot(df, aes(x = likes_count, y = comments_count))
p + geom_point() + scale_x_log10(labels = comma) + scale_y_log10(labels = comma) + 
        facet_wrap(~post_type, nrow = 2) 


# geom_text() allows to use party names instead of points
p <- ggplot(df[df$likes_count>10000, ], 
            aes(x = likes_count, y = comments_count, label = party))
p + geom_text() + scale_x_log10(labels = comma) + scale_y_log10(labels = comma)
```

Other examples:

```{r}
## Scatter plot with dots colored by type of post
p <- ggplot(df[df$likes_count>5000, ], 
            aes(x = likes_count, y = comments_count))
p + geom_point(aes(color = post_type)) + scale_x_log10(labels = comma) +
  scale_y_log10(labels = comma)

## Same for point shape
p <- ggplot(df[df$likes_count>5000,], 
            aes(x = likes_count, y=  comments_count))
p + geom_point(aes(shape = post_type)) + scale_x_log10(labels = comma) +
  scale_y_log10(labels = comma)

## Combining both (now different shapes also have different colors)
p <- ggplot(df[df$likes_count>5000,], 
            aes(x = likes_count, y = comments_count))
p + geom_point(aes(shape = post_type, color = post_type)) +
  scale_x_log10(labels = comma) + scale_y_log10(labels = comma)
```

## Dealing with cases where a lot of points are in some areas

Jittering points can avoid "overplotting", however, can also easily be misleading:

```{r}
p <- ggplot(df, aes(x = party, y = comments_count))
p + geom_point()

# vs
p <- ggplot(df, aes(x = party, y = comments_count))
p + geom_jitter(position = position_jitter(width = .1, height=.1))
```

```{r}
# Baseline
p <- ggplot(df, aes(x=likes_count, y = comments_count))
p + geom_point() + scale_x_log10(labels = comma) + scale_y_log10(labels = comma) 

## Jittering points (yet, really changes the plot's appearance here and can be misleading)
p <- ggplot(df, aes(x = likes_count, y = comments_count))
p + geom_jitter(position = position_jitter(width = .5, height =.5)) + 
    scale_x_log10(labels = comma) + scale_y_log10(labels = comma)

## Transparency
p <- ggplot(df, aes(x = likes_count, y = comments_count))
p + geom_jitter(position = position_jitter(width = .5, height = .5), alpha = 1/25) + 
    scale_x_log10(labels = comma) + scale_y_log10(labels = comma)

## Hexbin (if error: run install.packages("hexbin"))
p <- ggplot(df[df$likes_count>0 & df$comments_count>0,], 
        aes(x=likes_count, y = comments_count))
p + geom_hex() + scale_x_log10(labels = comma) + scale_y_log10(labels = comma) + 
    scale_fill_continuous(trans="log10")

# Generally, plotting binned means (either over deciles or over fixed (log) intervals like here)
# can help with plots that contain a lot of points/observations.
# Furthermore, geom_rug() is another option to indicate where most mass is
p <- ggplot(df, aes(x = likes_count, y = comments_count))
p + geom_point() + geom_rug(color = "grey", alpha = 0.6) +
   stat_summary_bin(fun = 'mean', bins = 20, color = 'green', size = 2, geom = 'point') +
  scale_x_log10(labels = comma) + scale_y_log10(labels = comma)

```
