---
title: "Twitter Rest API"
author: "Pablo Barberá, Ken Benoit, Friedrich Geiecke"
date: "15 November 2022"
output: html_document
---

In this file, we will look at static Twitter data, either recent tweets or user-level information. This type of data can be retrieved with Twitter's REST API. We will use the `rtweet` package here.

Loading packages:

```{r}
library("rtweet")
library("tidyverse")
```

### Authenticating

First we need to authenticate. One approach which does not depict the token in the code, is to first store it in a list with the `rtweet_app()` function and then save it on your computer as an .rda file. Run the following code chunk only once:

```{r, eva = FALSE}
# Only necessary to run once
auth <- rtweet_app()
save(auth, file = "myauthentication.rda")
```

The token is now saved in `myauthentication.rda`. To load your authentication whenever you run a script with `rtweet` in the future, you can simply add and run the following code chunk in the beginning (it works if the `myauthentication.rda` file is in the working directly; if it is stored in another directory, e.g. outside a repo to avoid uploading the token information to GitHub, you can add a path to the file such as `load("some/path/to/myauthentication.rda")`):

```{r}
# Load the file containing the list with the token
load("myauthentication.rda")

# Set authentication as default for remainder of session
auth_as(auth)
```

Lastly, if the chunk below outputs `LSEnews` after running, we are good to go:

```{r}
lookup_users("LSEnews")$screen_name
```

### Searching recent tweets

Without academic research access, it is possible to download recent tweets, but only up those less than 7 days old, and in some cases not all of them.

The following code will get the last 1,000 tweets mentioning "brexit" and "fuel":

```{r}
tweets <- search_tweets("brexit AND fuel", n = 1000)
head(tweets)
```

What are the most popular hashtags?

```{r}
ht <- str_extract_all(tweets$text, "#[A-Za-z0-9_]+")
ht <- unlist(ht)
head(sort(table(ht), decreasing = TRUE))
```

You can check the documentation about the options for string search [here](https://dev.twitter.com/rest/public/search).

### Extracting users' profile information

This is how you would extract information from user profiles:

```{r}
users <- lookup_users(c("JoeBiden", "POTUS", "VP", "FLOTUS", "BarackObama"))
users
```

Which of these has the most followers?

```{r}
users %>% select(screen_name, followers_count) %>% arrange(desc(followers_count))
```

### Downloading recent tweets from a specific user

Download recent tweets from a specific account (the maximum number of most recent tweets that can be freely downloaded per account is ~3200):

```{r}
tweets <- get_timeline("JoeBiden", n = 1000)
```

What are the accounts most commonly mentioned?

```{r}
ht <- str_extract_all(tweets$text, "@[A-Za-z0-9_]+")
ht <- unlist(ht)
head(sort(table(ht), decreasing = TRUE))
```

### Other types of data

The REST API also offers a long list of other endpoints that could be of use at some point, depending on your research interests. For example, if you know the ID of the tweets, you can download them directly from the API. This is useful because tweets cannot be redistributed as part of the replication materials of a published paper, but the list of tweet IDs can sometimes be shared. For example, a recent tweet of Barack Obama:

```{r}
# Downloading tweets when you know the ID
tw <- lookup_tweets(1320477828626948096)
tw$text
```

Furthermore, lists of Twitter users, compiled by other users, can also be accessed through the API. For example, this obtains a list of 550 Twitter accounts of US members of congress:

```{r}
# Download user information from a list (words in the list name are separated
# by a hyphen when using it as input into the slug variable)
congress_members <- lists_members(slug = "members-of-congress", owner_user = "cspan", token = auth)
head(congress_members)
```

Lastly, note that academic research track access to the Rest API is much more extensive. You can apply [here](https://developer.twitter.com/en/products/twitter-api/academic-research) and also use the `academicTwitteR` package.
