---
title: "MY472 - Week 7: A first script using RSelenium"
author: "Friedrich Geiecke"
date: "8 November 2022"
output: html_document
---

Note: Using `RSelenium` usually requires Java DK. First, try to see whether it is already installed on your computer - only install `RSelenium` with `install.packages("RSelenium")` and try to run the code in this document. If that does not work properly, next try to install Java DK. You can download the current version from here: https://www.oracle.com/java/technologies/downloads/. After its installation, restart RStudio.

Loading the Selenium package:

```{r}
#install.packages("RSelenium") # run once to install the package on your computer
library("RSelenium")
```

Launching the driver and browser (if the port is already in use, choose a different number with four digits, e.g. `rsDriver(browser=c("firefox"), port = 1234L)`:

```{r}
rD<- rsDriver(browser=c("firefox"))
driver <- rD$client
```

Navigate to the Google website:

```{r}
url <- "https://www.google.com/"
driver$navigate(url)
```

Since recently there can be new privacy terms that have to be accepted once in some regions.

Side note: Some such cases with pop-up windows on websites might require frame switching (although it seems that the 2022 version of Google does not require this at the moment). Should the code below not run on your computer after you have entered the correct XPath, try to uncomment the `swithToFrame` functions. As the the window is in the foreground, it can be that we have to switch the frame. Without this switching of the frame, we might not be able to click on the right element. Whether switching the frame is necessary depends on the design of the underlying website which can change.

As an exercise, the XPaths of the relevant elements have to be obtained with the Inspect function of the browser and then be pasted into this code replacing the "tba".

```{r}
#driver$switchToFrame(0) # can be un-commented and tried if code does not run

agree_button <- driver$findElement(using = "xpath", value = "tba")
agree_button$clickElement()

#driver$switchToFrame(1)
```

Next, we will search for the LSE:

```{r}
search_field <- driver$findElement(using = "xpath", value = "tba")
search_field$sendKeysToElement(list("london school of economics"))
Sys.sleep(1)
search_field$sendKeysToElement(list(key = "enter"))
```

And navigate to its website by clicking on the first link of the search results:

```{r}
first_link <- driver$findElement(using = "xpath", value = "tba")
first_link$clickElement()
```

Lastly, let us close the driver and browser:

```{r}
driver$close()
rD$server$stop()
```

