---
title: "MY472 - Week 7: A first script using RSelenium"
date: "Autumn Term 2024"
output: html_document
---

```{r setup, include=FALSE} 
# Because this script creates a marionette browser using RSelenium, I want to 
# be able to knit to HTML without evaluating each chunk.
knitr::opts_chunk$set(eval=FALSE)
```

**Important set-up steps:**

1. Using `RSelenium` usually requires Java DK. First, try to see whether it is already installed on your computer - only install `RSelenium` with `install.packages("RSelenium")` and try to run the code in this document. If that does not work properly, next try to install Java DK. You can download the current version from here: <https://www.oracle.com/java/technologies/downloads/>. After its installation, restart RStudio.

2. We will use the Firefox browser with `RSelenium` because it tends to work best. If it is not already installed on your computer, please install it from here: <https://www.mozilla.org/firefox/new/>.

First, let's load the `RSelenium` package as well as the `netstat` package. If they are not installed, then you can run the two `install.packages` command below.

```{r}
#install.packages("RSelenium") -- if needed, run once to install the package on your computer
#install.packages("netstat") -- if needed, run once to install the package on your computer
library(RSelenium)
library(netstat) # this is optional, to allow the use of free_port()
```

Now, we will launching the driver and browser (if the port is already in use, choose a different number with four digits, e.g. `rsDriver(browser=c("firefox"), port = 1234L)`). Alternatively -- as we do here -- choose a random free port using `netstat:free_port`. This code will now open what we call a 'marionette' browser (you can figure out why). Do not close this browser window!

```{r}
rD <- rsDriver(browser=c("firefox"), port = free_port(random = TRUE))
driver <- rD$client
```

Getting `RSelenium` to work might take a little effort. For example, some users will get the following error message when they try to launch the browser:

```
Error in if (file.access(phantompath, 1) < 0) { : 
  argument is of length zero
```

If you get this error, you can try to run these lines in R and then restart RStudio.

```{r, eval=FALSE}
# binman::rm_platform("phantomjs")
# wdman::selenium(retcommand = TRUE)
```

See this page for more information: <https://stackoverflow.com/questions/46202062/i-got-error-like-error-in-if-file-accessphantompath-1-0-argument-is-o> about this specific error. More generally, you should use Google to find resources to help you troubleshoot any other errors you get. In case you haven't noticed this yet: Stack Overflow is an amazing crowd-sourced resource for fixing coding errors.

Now, let's navigate to the Google website:

```{r}
url <- "https://www.google.com/"
driver$navigate(url)
```

These days, many countries and regions of the world require websites to allow users to opt out of accepting cookies. When you navigate to Google, you may get a pop-up window asking you to accept or reject. 

Side note: some websites with pop-up windows might require "frame switching" in order to click the accept or reject button. As of November 2024, it does not appear that Google requires this. However, if you find that your code does not work after finding the accept/reject buttons using XPath, then you should try: switch frames using `driver$switchToFrame(0)`, click the right button, then switch frames again using `driver$switchToFrame(1)`. Whether switching the frame is necessary depends on the design of the underlying website which can change.

Below, you should choose whether you want to accept or reject the cookies (your choice!). I will reject them by using XPath to find the element of the page with the text "Reject all".

```{r}
# driver$switchToFrame(0) # can be un-commented and tried if code does not run

# agree_button <- driver$findElement(using = "xpath", value = '//div[text()="Accept all"]') # notice that the "Accept all" text is in a div tag!
# agree_button$clickElement()
reject_button <- driver$findElement(using = "xpath", value = "//div[text()='Reject all']") # notice that the "Reject all" text is in a div tag!
reject_button$clickElement()

# driver$switchToFrame(1) # can be un-commented and tried if code does not run
```

Notice the use of single and double quotes in the XPaths above. If you use double quotes inside the XPath, you need to use single quotes around the whole XPath (and vice versa). This is an important issue in coding with strings.

Next, we will search for the LSE. Important note: you will very often need `Sys.sleep()` with `RSelenium`, because it runs too fast for many web browsers and/or webservers. In fact, we will consider it good etiquette to _always_ include delays when you are interacting with a webpage (i.e., making requests to a web server).

```{r}
search_field <- driver$findElement(using = "xpath", value = 'tbd')
search_field$sendKeysToElement(list("london school of economics"))
Sys.sleep(1) 
search_field$sendKeysToElement(list(key = "enter"))
```

And navigate to the LSE website by clicking on the first link of the search results:

```{r}
first_link <- driver$findElement(using = "xpath", value = 'tbd')
first_link$clickElement()
```

Lastly, let us close the driver and browser. Check that your marionette browser is actually closed now.

```{r}
# close the Rselenium processes:
driver$close()
rD$server$stop()

# close the associated Java processes (if using Mac or Linux this may not be necessary -- Google for correct command)
system("taskkill /im java.exe /f", intern=FALSE, ignore.stdout=FALSE)
```

