# Creating and querying a SQL database

Practice creating a SQL database and querying it. You should use the data in the file UK-tweets.csv. These data represent a random sample of 10,000 tweets related to the 2014 EP elections in the UK.

1. Create a SQLite database from the data in the file. Test the database by checking the first 5 entries in it.

```{r}
library(DBI)
db <- dbConnect(RSQLite::SQLite(), "data/tweet-db.sqlite")

tweets <- read.csv("UK-tweets.csv")
dbWriteTable(db, "tweets", tweets)

dbListFields(db, "tweets")
dbGetQuery(db, 'SELECT * FROM tweets LIMIT 5')
```

2. How many tweets are classified as engaging? (HINT: Use CASE WHEN)

```{r}
dbGetQuery(db, 
  "SELECT SUM(CASE WHEN communication='engaging' THEN 1 ELSE 0 END)  
    AS number_engaging FROM tweets")
```

3. What perecent of tweets are classified as impolite?

```{r}
# we need to multiply by a float (100.0) because otherwise 
# SQL will return only the integer part
dbGetQuery(db, 
  "SELECT (100.0 * SUM(CASE WHEN polite='impolite' THEN 1 ELSE 0 END) / COUNT(*)) 
    AS percent_impolite FROM tweets")
```

4. What percent of tweets mention Brexit?

```{r}
dbGetQuery(db, 
  "SELECT (100.0 * SUM(CASE WHEN text LIKE '%brexit%' THEN 1 ELSE 0 END) / COUNT(*)) 
    AS percent_brexit FROM tweets")
```


5. Compute the average number of followers for tweets that are impolite and polite.

```{r}
# again, multiply by 1.0 before dividing if you'd like to get decimal points
dbGetQuery(db, 
  "SELECT polite, (1.0 * SUM(followers_count) / COUNT(*)) 
    AS avg_followers 
    FROM tweets
    GROUP BY polite")
```

